---
title: "01 — Behavioural Analyses"
author: "[Author]"
date: today
format: html
---

## Overview

This document presents descriptive and exploratory behavioural analyses of
human performance in the Inspection Game. Participants (employees) played 150
trials against a computer-controlled employer, deciding to **work** or
**shirk** each round.

Two analyses are reported:

1. **Summary statistics** — demographic and trial-level descriptives across
   the full cleaned sample.
2. **Payoff by age group and gender** — mean payoff per trial as a function
   of age group (young vs. old) and gender.

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(janitor)
library(patchwork)
library(scales)
library(knitr)
library(kableExtra)
library(gt)

# Ensure output directories exist
dir.create(here("outputs", "figures"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("outputs", "tables"),  showWarnings = FALSE, recursive = TRUE)

# Load cleaned dataset produced by r/00_data_preparation.R
clean <- read_csv(
  here("data", "processed", "ig_data_clean.csv"),
  col_types = cols(
    id           = col_integer(),
    group        = col_integer(),
    choice       = col_integer(),
    opchoice     = col_integer(),
    payoff       = col_double(),
    trial        = col_integer(),
    age          = col_double(),
    gender       = col_integer(),
    group_label  = readr::col_factor(levels = c("young", "old")),
    gender_label = readr::col_factor(levels = c("male", "female"))
  )
)
```

---

## Analysis 1 — Summary Statistics

Descriptive statistics for age, gender, and valid trial counts across the
cleaned sample.

```{r summary-stats}
# ── Participant-level summary ────────────────────────────────────────────────
participant_summary <- clean |>
  group_by(id) |>
  summarise(
    age          = first(age),
    group_label  = first(group_label),
    gender_label = first(gender_label),
    n_trials     = n(),
    .groups = "drop"
  )

# ── Age: continuous summary ──────────────────────────────────────────────────
age_summary <- participant_summary |>
  group_by(group_label) |>
  summarise(
    n      = n(),
    mean   = mean(age),
    sd     = sd(age),
    min    = min(age),
    max    = max(age),
    .groups = "drop"
  )

# ── Gender counts ────────────────────────────────────────────────────────────
gender_summary <- participant_summary |>
  count(group_label, gender_label) |>
  pivot_wider(names_from = gender_label, values_from = n, values_fill = 0)

# ── Trial count summary (should all be 150 post-exclusion) ──────────────────
trial_summary <- participant_summary |>
  summarise(
    min_trials  = min(n_trials),
    max_trials  = max(n_trials),
    mean_trials = mean(n_trials)
  )

cat("Trial counts after exclusion — min:", trial_summary$min_trials,
    "max:", trial_summary$max_trials, "\n")

# ── Print tables ─────────────────────────────────────────────────────────────
age_summary |>
  gt() |>
  tab_header(title = "Age Summary by Group") |>
  fmt_number(columns = c(mean, sd), decimals = 1) |>
  cols_label(
    group_label = "Group",
    n           = "N",
    mean        = "Mean Age",
    sd          = "SD",
    min         = "Min",
    max         = "Max"
  )
```

```{r summary-gender-table}
gender_summary |>
  gt() |>
  tab_header(title = "Gender Counts by Group") |>
  cols_label(group_label = "Group")
```

```{r summary-plots}
# ── Figure 1a: Age distribution by group ─────────────────────────────────────
p_age <- ggplot(participant_summary, aes(x = age, fill = group_label)) +
  geom_histogram(binwidth = 2, colour = "white", alpha = 0.85, position = "identity") +
  facet_wrap(~group_label, ncol = 1) +
  scale_fill_manual(values = c("young" = "#4C72B0", "old" = "#DD8452")) +
  labs(
    title   = "Age Distribution by Group",
    x       = "Age (years)",
    y       = "Count",
    fill    = "Group"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

# ── Figure 1b: Gender counts by group ────────────────────────────────────────
gender_counts <- participant_summary |>
  count(group_label, gender_label)

p_gender <- ggplot(gender_counts, aes(x = group_label, y = n, fill = gender_label)) +
  geom_col(position = "dodge", colour = "white", alpha = 0.85) +
  scale_fill_manual(values = c("male" = "#55A868", "female" = "#C44E52")) +
  labs(
    title = "Gender Counts by Group",
    x     = "Group",
    y     = "Count",
    fill  = "Gender"
  ) +
  theme_bw(base_size = 12)

# ── Combine and save ──────────────────────────────────────────────────────────
fig1 <- p_age | p_gender
fig1

ggsave(
  here("outputs", "figures", "fig1_descriptives.png"),
  plot   = fig1,
  width  = 10,
  height = 5,
  dpi    = 300
)

# ── Save age summary table ────────────────────────────────────────────────────
write_csv(age_summary, here("outputs", "tables", "tbl1_age_summary.csv"))
write_csv(gender_counts, here("outputs", "tables", "tbl2_gender_counts.csv"))
```

---

## Analysis 2 — Mean Payoff by Age Group and Gender

Mean payoff per trial as a function of age group and gender. Missing trials
are excluded from this analysis (i.e., only rows with a non-missing `payoff`
value are used). Payoff is summarised at the participant level first (mean
payoff per trial), then plotted across groups.

```{r payoff-prep}
# ── Participant-level mean payoff (excluding any missing payoff values) ───────
payoff_by_participant <- clean |>
  filter(!is.na(payoff)) |>
  group_by(id, group_label, gender_label) |>
  summarise(
    mean_payoff = mean(payoff),
    .groups = "drop"
  )

# ── Group-level summary for annotations ──────────────────────────────────────
payoff_group_summary <- payoff_by_participant |>
  group_by(group_label, gender_label) |>
  summarise(
    n           = n(),
    mean        = mean(mean_payoff),
    se          = sd(mean_payoff) / sqrt(n()),
    .groups = "drop"
  )

payoff_group_summary |>
  gt() |>
  tab_header(title = "Mean Payoff per Trial by Group and Gender") |>
  fmt_number(columns = c(mean, se), decimals = 2) |>
  cols_label(
    group_label  = "Group",
    gender_label = "Gender",
    n            = "N",
    mean         = "Mean Payoff",
    se           = "SE"
  )
```

```{r payoff-plots}
# ── Figure 2a: Payoff by age group ───────────────────────────────────────────
p_payoff_group <- ggplot(payoff_by_participant,
                          aes(x = group_label, y = mean_payoff, fill = group_label)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.06, size = 1.2, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", colour = "black") +
  scale_fill_manual(values = c("young" = "#4C72B0", "old" = "#DD8452")) +
  labs(
    title = "Mean Payoff per Trial by Age Group",
    x     = "Group",
    y     = "Mean Payoff (cents)",
    fill  = "Group"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

# ── Figure 2b: Payoff by gender ───────────────────────────────────────────────
p_payoff_gender <- ggplot(payoff_by_participant,
                           aes(x = gender_label, y = mean_payoff, fill = gender_label)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.06, size = 1.2, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", colour = "black") +
  scale_fill_manual(values = c("male" = "#55A868", "female" = "#C44E52")) +
  labs(
    title = "Mean Payoff per Trial by Gender",
    x     = "Gender",
    y     = "Mean Payoff (cents)",
    fill  = "Gender"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

# ── Figure 2c: Interaction — group × gender ───────────────────────────────────
p_payoff_interaction <- ggplot(payoff_group_summary,
                                aes(x = group_label, y = mean,
                                    colour = gender_label, group = gender_label)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.1) +
  scale_colour_manual(values = c("male" = "#55A868", "female" = "#C44E52")) +
  labs(
    title  = "Mean Payoff: Group × Gender Interaction",
    x      = "Group",
    y      = "Mean Payoff ± SE (cents)",
    colour = "Gender"
  ) +
  theme_bw(base_size = 12)

# ── Combine and save ──────────────────────────────────────────────────────────
fig2 <- (p_payoff_group | p_payoff_gender) / p_payoff_interaction
fig2

ggsave(
  here("outputs", "figures", "fig2_payoff_by_group_gender.png"),
  plot   = fig2,
  width  = 10,
  height = 8,
  dpi    = 300
)

# ── Save payoff summary table ─────────────────────────────────────────────────
write_csv(payoff_group_summary, here("outputs", "tables", "tbl3_payoff_summary.csv"))
```
